### 2019-12-15 ic-journal/multivariate.tex
text

\hl{In this section we attack the case of 
  multiple equations (the ``\multivariate''  case), for two major theorems earlier discussed:}
Milner's ``unique solution of equations (for $\sim$)''
(Theorem~\ref{t:Mil89s1}) and  the  ``unique solution of rooted
contractions'' (Theorem~\ref{t:rcontraBisimulationU}).
%
The formalisation supports finitely many
  equations/contractions and equation variables.\footnote{The original theorems
hold for even infinitely many equations and equation variables.}
We chose these two theorems because \hlD{of their relevance}, and 
because they well illustrate the \hlD{effort} needed in the \multivariate case.
% The work is a \hl{natural} extension of the existing work, for
% \hl{no new algebraic type (e.g. the type for CCS processes) has to be defined,}
% and neither the CCS syntax nor the SOS rules \hl{needs} to be changed.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}) and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in \hl{the finite domain of} \HOLinline{\HOLFreeVar{fm}} are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. \hl{Initially,} such a finite map can be built from
a list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding substituted terms
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
omitted here).
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is usually
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
The substitution mechanism is order-independent, \hlD{as based
 maps.} For most CCS operators it
just recursively calls itself on subterms.  \hlD{The} most
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

@@

It could be possible to inductively define \multivariate contexts and their
guarded variants, but care is needed
for 
 variables that occur \hl{within a  recusion}.  \hl{We have preferred a different
   solution,} in which the definitions of 
 \multivariate contexts  is based on the
 existing \univariate definitions.
\hlD{For this, intuitively,  we replace a single variable with a hole and (viewing
the other variables as constants) so to obtain a  single-variable context
on which predicates like} \HOLinline{\HOLConst{CONTEXT}} can be used.
For example, to see the weak guardedness of $a.X + b.X + c.Y$, 
we focus on each (free) variable: if we substitute each
occurrence of a variable, say $X$, with a ``hole'' and see the
resulting term as a $\lambda$-function, then all such $\lambda$-functions
must satisfy \HOLinline{\HOLConst{WG}}, i.e.~the following results hold:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{Y}\ensuremath{)}
\HOLTokenTurnstile{} \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{X} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{X} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}
\end{alltt}
Note that \HOLinline{\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLFreeVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLFreeVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{Y}\\\;\HOLSymConst{\ensuremath{=}}\\\;\ensuremath{[}\HOLFreeVar{t}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\ensuremath{(}\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{Y}\ensuremath{)}}
can be expressed as a \univariate substitution of the original term.
This idea leads to the following definitions:
\begin{alltt}
   \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{X}. \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \ensuremath{[}\HOLBoundVar{t}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Xs}\hfill{[context_def]}

@@

For weakly guarded contexts (which are also normal contexts), beside
their regular properties (i.e. weak
guardedness) as in the \univariate case (\HOLinline{\HOLConst{WG}}), we also need
their composability w.r.t.~\multivariate contexts: if we
substitute some free variables in a context $C$ with some weakly guarded
contexts, the resulting \hlD{context is} weakly guarded (w.r.t.~the same set of variables):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}\ensuremath{)}\hfill{[weakly_guarded_combin]}
\end{alltt}

@@

Now consider (formally) what is a solution $\til P$ of CCS equations. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that a single definition
can be used to represent solutions of all kinds of CCS
\hl{equations and inequations.} Then, of course ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$ must hold:
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \cdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be a pure process, i.e.~having no free
variable. Finally, the set of bound variables is disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs
easier.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes. Furthermore, \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}} implicitly
  implies that the two lists $A$ and $B$ have the same length.}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}
Note that the two logical terms ``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}\\\;\HOLFreeVar{Ps}}'' and ``\HOLinline{\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenIn{}}\\\;\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' in
HOL have the same meaning. The latter form suggests that,
``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' is actually a set containing all
solutions of ``\HOLinline{\HOLConst{CCS_equation}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}''. Then the unique-solution
theorems can be understood thus: any two (syntactically different) elements
in the solution set are bisimilar.

>>>>

\hl{In this section we attack the case of 
  multiple equations (the `\multivariate'  case), for two major theorems earlier discussed:}
Milner's ``unique solution of equations (for $\sim$)''
(Theorem~\ref{t:Mil89s1}) and  the  ``unique solution of rooted
contractions'' (Theorem~\ref{t:rcontraBisimulationU}).
%
The formalisation supports finitely many
  equations/contractions and equation variables.\footnote{The original theorems
hold for even infinitely many equations and equation variables.}
We chose these two theorems because \hlD{of their relevance}, and 
because they well illustrate the \hlD{work} needed in the \multivariate case.
% The work is a \hl{natural} extension of the existing work, for
% \hl{no new algebraic type (e.g. the type for CCS processes) has to be defined,}
% and neither the CCS syntax nor the SOS rules \hl{needs} to be changed.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}) and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in \hl{the finite domain of} \HOLinline{\HOLFreeVar{fm}} are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. \hl{Initially,} such a finite map can be built from
a list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding substituted terms
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
omitted here).
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is usually
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
Using maps, the substitution mechanism is order-independent. For most CCS operators,
\HOLinline{\HOLConst{CCS_SUBST}}
 recursively calls itself on subterms.  \hlD{The} most
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

@@

It could be possible to inductively define \multivariate contexts and their
guarded variants, but care is needed
for 
 variables that occur \hl{within a  recusion}.  \hl{We have preferred a different
   solution,} in which the definitions of 
 \multivariate contexts  is based on the
 existing \univariate definitions.
\hlD{For this, intuitively,  we replace a single variable with a hole  (viewing
the other variables as constants), so to obtain a  single-variable context
on which predicates like} \HOLinline{\HOLConst{CONTEXT}} can be used.
For example, to see the weak guardedness of $a.X + b.X + c.Y$, 
we focus on each (free) variable, we substitute each
occurrence of a variable, say $X$, with a ``hole'' and see the
resulting term as a $\lambda$-function; then all such $\lambda$-functions
should satisfy \HOLinline{\HOLConst{WG}}, i.e.~the following results hold:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{Y}\ensuremath{)}
\HOLTokenTurnstile{} \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{X} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var} \HOLFreeVar{X} \HOLSymConst{\ensuremath{+}} \HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}
\end{alltt}
Note that \HOLinline{\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLFreeVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLFreeVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{Y}\\\;\HOLSymConst{\ensuremath{=}}\\\;\ensuremath{[}\HOLFreeVar{t}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\ensuremath{(}\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{b}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{c}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{Y}\ensuremath{)}}
can be expressed as a \univariate substitution of the original term.
This idea leads to the following definitions:
\begin{alltt}
   \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{X}. \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \ensuremath{[}\HOLBoundVar{t}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Xs}\hfill{[context_def]}

@@

For weakly guarded contexts, % (which are also normal contexts), 
besides
their usual properties  as in the \univariate case (i.e., weak
guardedness,
\HOLinline{\HOLConst{WG}}), we also need
their composability w.r.t.~\multivariate contexts: if we
substitute some free variables in a context $C$ with some weakly guarded
contexts, the resulting \hlD{context is} weakly guarded (w.r.t.~the same set of variables):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}\ensuremath{)}\hfill{[weakly_guarded_combin]}
\end{alltt}

@@

Now consider (formally) what is a solution $\til P$ of CCS equations. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that a single definition
can be used to represent solutions of all kinds of CCS
\hlD{equations.} Then,  ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$ must hold:
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \cdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be a pure process, i.e.~having no free
variable. Finally, the set of bound variables is disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs
easier.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes with the same length.
 % Furthermore,  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}}
  implicitly   implies that the two lists $A$ and $B$ have the same length.
}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}
Note that the two logical terms ``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}\\\;\HOLFreeVar{Ps}}'' and ``\HOLinline{\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenIn{}}\\\;\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' in
HOL have the same meaning. The latter form suggests that,
``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' is actually a set containing all
solutions of ``\HOLinline{\HOLConst{CCS_equation}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}''. Then the unique-solution
theorems can be understood thus: any two (syntactically different) elements
in the solution set are bisimilar.
